---
title: "LandfillShrubs.qmd"
format: html
editor: visual
---

## Import Data

### Import library

```{r}
library(readr)
library(assertthat)
library(tidyverse)
library(lmerTest)
library(lme4)
library(vegan)
```

### Shrub data

```{r}
data.march <- read.csv("./raw-data/Landfill_shrub_March_data.csv")
data.april <- read.csv("./raw-data/Landfill_shrub_data.csv")

# drop rows without a date, since there are lots of rows with empty strings
data.march[data.march==''] = NA
data.march = data.march |> filter(!is.na(Date))
data.april[data.april==''] = NA
data.april = data.april |> filter(!is.na(Date))

## Drop first row, since it is descriptive
data.april = data.april[2: nrow(data.april),]

# Create shrub ids
data.april$shrub_id <- paste(data.april$Site, data.april$Shrub_location, sep='-')
data.march$shrub_id <- paste(data.march$Site, data.march$Shrub_location, sep='-')
```

All Plant IDs

-   There are some plants only from march, but every april plant was also in march

```{r}
assert_that(sum(!(data.april$shrub_id %in% data.march$shrub_id)) == 0,
            msg = "All april shrubs should be present in march data")
```

### Soil Data

```{r}
soil_raw = readxl::read_excel("./raw-data/soil.XLSX")

# observation rows
column_names = unname(c(soil_raw[8,1:2], soil_raw[7,3:28])) |>
  str_replace(' ', "_") |> 
  str_replace("0\\.", "pt") |> 
  str_replace('\\)', '') |> 
  str_replace('\\(','') |>
  str_replace('\\r\\n', "_")

units = unname(c(soil_raw[8,1:28])) |> 
  str_extract("(?<=\r\n).*") |> 
  str_replace("/", "_")
units[1:2] = ""

column_names = paste(column_names, units, sep="_")
x = bind_rows(soil_raw[9:14,28], soil_raw[25:39,])

soil1 = soil_raw[9:14,1:28]
colnames(soil1) = column_names
soil2 = soil_raw[25:39,1:23]
colnames(soil2) = column_names

soil_messy = bind_rows(soil1, soil2) 

soil_messy = soil_messy |> rename(
  Lab_sample_num = `LAB_SAMPLE_#_`,
  Client_description = Client__Description_,
  pH = pH_,
  SAR = SAR_
)
```

## Tidy format

```{r}
data = bind_rows(
  data.march |> 
    rename(
      Height_cm = Height_cm_1,
      Plant_status = Plant_status_1
      ) |> 
    mutate(
      Shrub_location = as.numeric(Shrub_location),
      Date = mdy(Date)
      ),
  data.april |> 
    rename(
      Height_cm = Height_cm_2,
      Plant_status = Plant_status_2
      ) |> 
    mutate(
      Shrub_location = as.numeric(Shrub_location),
      Height_cm = as.numeric(Height_cm),
      Date = mdy('4/30/2024')
      ))
```

### Soil

```{r}
soil_messy[soil_messy == "ISM"] = NA

soil = soil_messy |>
   fill(
     Client_description,
     `SP_%`,
     .direction = "down") |> 
   separate(Lab_sample_num, into = c('Sample_num', 'Lab_sample_test_num'), fill = 'right') |>
   mutate(
     Lab_sample_test_num = case_when(
       Lab_sample_test_num == "dup" ~ 2,
       is.na(Lab_sample_test_num) ~ 1
       )
     ) |>
   separate(Client_description, into = c('Location', 'Depth')) |> 
   mutate_all(function(x) str_replace(x, "<", ''))
```

## Validation

We would expect that:

-   Plants with same shrub_id have same species

    ```{r}
    # Resolution

    mismatching_species = data |> 
      select(c(shrub_id, Species, Date, Plant_status, Notes)) |> 
      pivot_wider(names_from = Date, values_from = c(Species, Plant_status, Notes)) |> 
      filter(`Species_2024-03-01` != `Species_2024-04-30`)

    # For the one that was first alive and is now dead, use the species from when it was alive
    data[data$shrub_id == "LF5S-14" & data$Date == "2024-04-30", "Species"] = 'ARTCAL'

    mismatching_species = data |> 
      select(c(shrub_id, Species, Date, Plant_status, Notes)) |> 
      pivot_wider(names_from = Date, values_from = c(Species, Plant_status, Notes)) |> 
      filter(`Species_2024-03-01` != `Species_2024-04-30`)

    # Filter the rest out
    data = data |> 
      filter(!(shrub_id %in% mismatching_species$shrub_id))

    mismatching_species = data |> 
      select(c(shrub_id, Species, Date, Plant_status, Notes)) |> 
      pivot_wider(names_from = Date, values_from = c(Species, Plant_status, Notes)) |> 
      filter(`Species_2024-03-01` != `Species_2024-04-30`)


    assert_that(
      nrow(mismatching_species) == 0,
      msg = "Should be no mismatching species"
    )

    # (`[[`)("colname")),
    ```

-   Plants would not be dead in march and alive in april

    ```{r}
    undead_shrubs = data |> 
      select(c(shrub_id, Date, Plant_status)) |> 
      pivot_wider(names_from = Date, values_from = Plant_status) |> 
      filter(`2024-03-01` == 'Dead') |> 
      filter(`2024-04-30` == 'Alive')

    assert_that(
      nrow(undead_shrubs) == 0,
      msg = "Should be no vampire shrubs that came back from the dead"
    )
    ```

-   Plants that have missing species in April are the same as in March

    ```{r}
    data = data |> 
      select(c(shrub_id, Date, Species)) |> 
      pivot_wider(
        values_from = Species,
        names_from = Date
      ) |> 
      mutate(
        Species = `2024-03-01`
      ) |> 
      select(-c(`2024-03-01`, `2024-04-30`)) |> 
      left_join(data |> select(-c(Species)), join_by(shrub_id))
    ```

-   Plants that have missing status data and were dead in march are still dead

    ```{r}
    data = data |> 
      select(c(shrub_id, Date, Plant_status)) |> 
      pivot_wider(values_from = Plant_status, names_from = Date) |> 
      mutate(
        `2024-04-30` = case_when(
          `2024-03-01` == "Dead" & is.na(`2024-04-30`) ~ "Dead",
          .default = `2024-04-30`
          ) 
        ) |>    
      pivot_longer(!shrub_id, names_to = "Date",  values_to = "Plant_status") |> 
      mutate(
        Date = ymd(Date)
      ) |> 
      right_join(
        data |> select(-c(Plant_status)), 
        by=join_by(Date, shrub_id))

    ```

## Cleaning

1.  Get rid of unnecessary columns

    ```{r}
    measurements = data |> select(-c(Team, Chlorophyll_Team)) |> mutate(
        Month = case_when(
          month(Date) == 3 ~ "March",
          month(Date) == 4 ~ "April")) |> mutate(
      Site_numeric_id = 
        as.numeric(factor(Site, levels=unique(Site)))
    )

    shrubs = measurements |> 
      pivot_wider(
        id_cols = c(shrub_id),
        names_from = c(Date),
        values_from = c(
          Height_cm, 
          Plant_status, 
          Notes,
          Leaf1_Chlorophyll, 
          Leaf2_Chlorophyll, 
          Leaf3_Chlorophyll, 
          Chlorophyll_Team_Notes,
          Month
          ),
        unused_fn = unique
    ) |> select(-c(
      `Leaf1_Chlorophyll_2024-03-01`, 
      `Leaf2_Chlorophyll_2024-03-01`, 
      `Leaf3_Chlorophyll_2024-03-01`, 
      `Chlorophyll_Team_Notes_2024-03-01`)
      ) |> rename(
      Leaf1_Chlorophyll = `Leaf1_Chlorophyll_2024-04-30`,
      Leaf2_Chlorophyll = `Leaf2_Chlorophyll_2024-04-30`,
      Leaf3_Chlorophyll = `Leaf3_Chlorophyll_2024-04-30`,
      Chlorophyll_Notes = `Chlorophyll_Team_Notes_2024-04-30` 
      ) |> mutate(
        Height_change_cm = `Height_cm_2024-04-30` - `Height_cm_2024-03-01`
      ) |> mutate(
        Percent_height_change = Height_change_cm / `Height_cm_2024-03-01`
      )
    ```

## Analysis

### Helpers

##### Plot Values

```{r}
orange_p = "#f2790f"
orange_s = "#f5cfae"

green_p = "#47805b"
green_s = "#8f8b89"

brown_p = "#785c44"
brown_s = "#bdb4ac"

scoured_primary = brown_p
scoured_secondary = brown_s
nonscoured_primary = orange_p
nonscoured_secondary = orange_s

soil_type_labels = c(N = "Unscoured", S = "Scoured")
species_labels = c(ENCCAL = "Encelia californica", ARTCAL = "Artemisia california", ERIFAS = "Eriogonum fasciculatum", SALMEL = "Salvia mellifera")
```

##### Write data

```{r}
SHOULD_WRITE_FILES = T
FIGURE_WIDTH = 7
FIGURE_HEIGHT = 4.5
OUTPUT_FILE_PATH = "./landfill-output/"

write_plot = function(output, plot_name, analysis_name) {
  if(SHOULD_WRITE_FILES) {
    ggsave(plot = output, width = FIGURE_WIDTH, height = FIGURE_HEIGHT,filename=paste(OUTPUT_FILE_PATH, analysis_name,"-",plot_name, ".jpg", sep = ""))
  }
}

write_table = function(model, model_name, analysis_name) {
  if(SHOULD_WRITE_FILES) {
    write.csv(model, paste(OUTPUT_FILE_PATH, analysis_name,"-",model_name, ".csv", sep = ""))
  }
}
```

### Summary

```{r}
measurements |> ggplot() + 
  aes(x = Species, fill = Species) + 
  geom_bar() + 
  facet_grid(~ Date)
  
measurements |> 
  filter(Plant_status == 'Alive') |> 
  ggplot() + 
  aes(x = Species, fill = Species) + 
  geom_bar() + 
  facet_grid(~ Date)

```

### Soil

```{r}

soil |> group_by(Location, Depth)

soil_matrix = as.matrix(mutate_all((soil[6:25]), as.numeric))
rownames(soil_matrix) = paste(soil$Location, soil$Depth, soil$Lab_sample_test_num, sep='-')


#prcomp(soil_matrix, scale = TRUE)
```

### Height Change

##### Data

-   Filter out data that don't have before and after heights

-   Create a set with only living shrubs, because if the plant is dead, its change in height may not be reflective.

-   [ ] should I try to treat status as a factor instead of filtering it out? come back to this.

```{r}
# shrubs with measurements at both sampling times
height_change_shrubs = shrubs |>
  filter(!is.na(Height_change_cm))

# shrubs with measurements at both sampling times, that are alive 
height_change_shrubs_alive = height_change_shrubs |>
  filter(`Plant_status_2024-04-30` == "Alive")
```

##### Visualize

-   Visualize

```{r}
# Shrub counts per species by Status and Soil Type
height_change_shrubs |> ggplot() + 
  aes(x = Species, fill = Species) + 
  geom_bar() + 
  facet_grid(`Plant_status_2024-04-30` ~ Soil) + 
  labs(title = "Shrub count by Status and Soil Type")

# Box / jitterplot per species by Status and Soil Type
height_change_shrubs |> ggplot() + 
  aes(x = Species, y = Height_change_cm, color = Soil) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(position = position_jitterdodge(jitter.width = 0.2), alpha = 0.8) + 
  facet_grid(~ `Plant_status_2024-04-30`)

# Box / jitterplot per species for living plants by Soil type
height_change_shrubs_alive |> ggplot() + 
  aes(x = Species, y = Height_change_cm, color = Soil) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(position = position_jitterdodge(jitter.width = 0.2), alpha = 0.8)

# Normalize height change using a percentage
height_change_shrubs_alive |> ggplot() + 
  aes(x = Species, y = Percent_height_change, color = Soil) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(position = position_jitterdodge(jitter.width = 0.2), alpha = 0.8)

# Bar graph with sd and error bars
height_change_shrubs_alive |> ggplot(
    aes(x = Species, y = Height_change_cm, fill = Species)
  ) + 
  stat_summary(
    fun = mean, 
    geom= "col", 
    position = "dodge"
  ) +
  stat_summary(
    fun.data = mean_se, 
    geom = "errorbar", 
    position = position_dodge(0.9), 
    width = 0.1
  ) 


# Bar graph with sd and error bars
height_box_plot = height_change_shrubs_alive |> ggplot(
    aes(x = Species, y = Percent_height_change, fill = Species)
  ) + 
  stat_summary(
    fun = mean, 
    geom= "col", 
    position = "dodge"
  ) +
  stat_summary(
    fun.data = mean_se, 
    geom = "errorbar", 
    position = position_dodge(0.9), 
    width = 0.1
  ) + 
  scale_fill_manual(
    labels = species_labels,
    values = c(viridis::viridis(5))
  ) +
  scale_x_discrete(
    labels = str_replace_all(species_labels, "\\s", "\n")
  ) +
  labs(
    title = "Height by Soil Type and Species",
    y = "Percent height change"
  ) +
  theme(
    axis.text.x = element_text(face = 'italic'),
    legend.text = element_text(face = 'italic')
    )

height_box_plot

write_plot(height_box_plot, "box-plot", "height")

# (`[[`)("station_code")
```

##### Models and Normality

```{r}
height_change_shrubs_alive |> ggplot() +
  aes(x = Height_change_cm) +
  geom_histogram()

height_model_interactions = 
  glm(Height_change_cm ~ Species * Soil, data = height_change_shrubs_alive)
height_model_no_interactions = 
  glm(Height_change_cm ~ Species + Soil, data = height_change_shrubs_alive)
height_model_species = 
  glm(Height_change_cm ~ Species, data = height_change_shrubs_alive)

plot(height_model_interactions)
plot(height_model_no_interactions)
plot(height_model_species) 

percent_height_model_interactions = glm(Percent_height_change ~ Species * Soil, data = height_change_shrubs_alive)
percent_height_model_no_interactions= glm(Percent_height_change ~ Species + Soil, data = height_change_shrubs_alive)
percent_height_model_species= glm(Percent_height_change ~ Species, data = height_change_shrubs_alive)
percent_height_model_soil= glm(Percent_height_change ~ Soil, data = height_change_shrubs_alive)


```

##### ANOVA

```{r}
AIC(height_model_interactions)
AIC(height_model_no_interactions)
AIC(height_model_species)

AIC(percent_height_model_interactions)
AIC(percent_height_model_no_interactions)

summary(height_model_interactions)
summary(height_model_no_interactions)
summary(height_model_species)

summary(percent_height_model_interactions)
summary(percent_height_model_no_interactions)
summary(percent_height_model_species)
summary(percent_height_model_soil)


```

###### Location as random effect?

```{r}
summary(lmer(Percent_height_change ~ Soil * Species + (1|Site_numeric_id), data = height_change_shrubs_alive |> mutate(
  Site_id = as.numeric(
    factor(Site, levels=unique(Site))
    )
  )))
```

##### Model Comparison

### Survival

#### Species and Soil

model: `Survival ~ Species * Soil`

##### Data

```{r}
survival_data = measurements |> 
  filter(Month == 'April') |> 
  select(c(shrub_id, Plant_status, Soil, Species, Notes, Site)) |>
  mutate(
    Survival_index = case_when(
      Plant_status == 'Alive' ~ 1, 
      Plant_status == 'Dead' ~ 0, 
      .default = NA
      )
    ) |>
  filter(!is.na(Survival_index))
```

##### Visualize

```{r}
survival_bar_plot = survival_data |> ggplot() + 
  aes(
    x = Soil, 
    fill = interaction(
      factor(Plant_status, levels = c("Dead", "Alive")),
      factor(Soil)
      ),
    ) +
  geom_bar(position = "stack") +
  facet_grid(~ Species) +
  scale_fill_manual(
    labels = c(
      Alive.S = "Alive (scoured)",
      Dead.S = "Dead (scoured)",
      Alive.N = "Alive (unscoured)",
      Dead.N = "Dead (unscoured)"
    ),
    values = c(
      Alive.S = scoured_primary, Dead.S = scoured_secondary,
      Alive.N = nonscoured_primary, Dead.N = nonscoured_secondary
      )
    ) + 
  labs(
    fill = "Soil type & Plant status",
    x = "Soil type",
    y = "Number of Plants",
    title = "Impact of soil type on plant survival per species"
    ) +
  scale_x_discrete(labels = soil_type_labels,
                   guide = guide_axis(angle = 60))

write_plot(
  survival_bar_plot,
  "bar-plot",
  "survival"
)

survival_bar_plot
```

##### Model and Normality

```{r}

soil_survival_model = glm(Survival_index ~ Soil * Species, data = survival_data, family = "binomial")

soil_survival_no_erifas_model = glm(Survival_index ~ Soil * Species, data = survival_data |> filter(
  Species != 'ERIFAS'
), family = "binomial")

plot(soil_survival_model)
plot(soil_survival_no_erifas_model)

```

Looks a bit better without "ERIFAS", which we will exclude since there are only 4 individuals total, and 2 per soil type.

##### Analysis and Tests

Chi square test

```{r}
survival_test = anova(soil_survival_model, test="Chisq")

write_table(survival_test, "chisq", "survival")

survival_test
```

Results

-   Significant(ish?) interaction between Soil and Species on survival, aka the survival of some species is more likely to be affected by soil type than others. However, take with a grain of salt as P value is `0.087`.

#### Species

```{r}
anova(glm(Survival_index ~ Soil, data = survival_data |> filter(
  Species == 'ENCCAL'
), family = "binomial"), test = "Chisq")

anova(glm(Survival_index ~ Soil, data = survival_data |> filter(
  Species == 'ARTCAL'
), family = "binomial"), test = "Chisq")

anova(glm(Survival_index ~ Soil, data = survival_data |> filter(
  Species == 'SALMEL'
), family = "binomial"), test = "Chisq")


```

#### Salinity

##### Data

```{r}
salinity_survival = soil |> 
  select(c(
    Location, 
    Depth, 
    EC_dS_m,
    Lab_sample_test_num
    )) |> 
  mutate(EC_dS_m= as.numeric(EC_dS_m)) |>
  inner_join(
    survival_data |> 
      select(c(
        shrub_id, 
        Site, 
        Survival_index,
        Species
        )),
    by = join_by(Location == Site)
  ) |>
  arrange(shrub_id) |> 
  mutate(
    Soil_id = paste(
      shrub_id,
      Depth,
      Lab_sample_test_num, 
      sep = "_")
  )


anova(glm( 
  Survival_index ~ EC_dS_m,
  data = salinity_survival,  
  family = binomial(link="logit") )
  )
```

##### Visualize

```{r}
salinity_survival |> filter(
    Lab_sample_test_num == 1
  )|> ggplot() +
  aes(x = EC_dS_m,y=Survival_index) +
  geom_point() +
  facet_grid(Species ~ Depth)
```

### Chlorophyll Content

##### Get data

```{r}
chlorophyll_data = measurements |> 
  filter(Month == 'April') |> 
  select(c(
    shrub_id, 
    Soil,
    Species,
    Leaf1_Chlorophyll, 
    Leaf2_Chlorophyll, 
    Leaf3_Chlorophyll, 
    Chlorophyll_Team_Notes
    )) |> 
  pivot_longer(
    cols = c(Leaf1_Chlorophyll, Leaf2_Chlorophyll, Leaf3_Chlorophyll),
    values_to = 'Chlorophyll',
    names_to = "Leaf"
  ) |> mutate(
    Leaf = case_when(
      Leaf == "Leaf1_Chlorophyll" ~ as.numeric(1),
      Leaf == "Leaf2_Chlorophyll" ~ as.numeric(2),
      Leaf == "Leaf3_Chlorophyll" ~ as.numeric(3)
    )
  ) |> filter(
    !is.na(Chlorophyll)
  ) 

 chlorophyll_data_mean = chlorophyll_data |> 
  group_by(shrub_id, Soil, Species, Chlorophyll_Team_Notes) |> 
   summarise(
    Mean_Chlorophyll = mean(Chlorophyll)
   ) |> ungroup()
```

##### Summarize Data

```{r}
chlorophyll_data
```

##### Visualize

```{r}
chlorophyll_data_mean |> ggplot() + 
  aes(x = Mean_Chlorophyll) +
  geom_histogram()

chlorophyll_barplot = chlorophyll_data_mean |> 
  complete(Soil, Species, fill = list(Mean_Chlorophyll = 0)) |>
  ggplot(
    aes(x = Species, y = Mean_Chlorophyll, fill = Soil)
  ) + 
  stat_summary(
    fun = mean, 
    geom= "col", 
    position = "dodge"
  ) +
  stat_summary(
    fun.data = mean_se, 
    geom = "errorbar", 
    position = position_dodge(0.9), 
    width = 0.1
  ) + 
  scale_fill_manual(values = c(N = nonscoured_primary, S = scoured_primary)) + 
  labs(
    title = "Impact of Soil on Chlorophyll Content per Species",
    y = "Mean Chlorophyll Content per Plant"
  )



chlorophyll_boxplot = chlorophyll_data_mean |>   
ggplot() + 
  aes(x = Soil, y = Mean_Chlorophyll, color = Soil, fill = Soil) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(
    shape=4,
    position = position_jitterdodge(jitter.width = 0.3), 
    alpha = 0.8) +
  scale_color_manual(
    values = c(N = nonscoured_primary, S = scoured_primary),
    labels = soil_type_labels
    ) +
  scale_fill_manual(
    values= c(N = nonscoured_secondary, S = scoured_secondary),
    labels=soil_type_labels
    ) + 
  facet_grid(~Species) +
  scale_x_discrete(
    labels =soil_type_labels,
    guide = guide_axis(angle = 60)
  ) +
  labs(
    title = "Impact of Soil on Chlorophyll Content per Species",
    y = "Mean Chlorophyll Content per Plant"
  )

write_plot(chlorophyll_barplot, "bar-plot", "chlorophyll")
write_plot(chlorophyll_boxplot, "box-plot", "chlorophyll")

chlorophyll_barplot
chlorophyll_boxplot
```

##### Models and Normality

```{r}
chlorophyll_with_interactions = glm(Mean_Chlorophyll ~ Species * Soil, data = chlorophyll_data_mean)

plot(chlorophyll_with_interactions)
```

Try to group by plant as a random effect

```{r}

model1 = lmer(Chlorophyll ~ Soil * Species + (1|shrub_id), data = chlorophyll_data)
```

##### ANOVA

```{r}

AIC(chlorophyll_with_interactions)
AIC(model1)

chlorophyll_test = summary(chlorophyll_with_interactions)

# AIC is much worse for this one even though the results are similar
summary(model1) 

write_table(chlorophyll_test$coefficients, "anova", "chlorophyll")

chlorophyll_test
chlorophyll_test$coefficients
```

##### Anovas for individual species

```{r}
enccal_chlorophyll_model = glm(Mean_Chlorophyll ~ Soil, data = chlorophyll_data_mean |> filter(
  Species == "ENCCAL"
)) 

get_df(enccal_chlorophyll_model)

t.test(Mean_Chlorophyll ~ Soil, data = chlorophyll_data_mean |> filter(Species == "ENCCAL"))
enccal_chlorophyll = enccal_chlorophyll_model |> summary()

enccal_chlorophyll

write_table(enccal_chlorophyll$coefficients, "anova-enccal", "chlorophyll")

artcal_chlorophyll = glm(Mean_Chlorophyll ~ Soil, data = chlorophyll_data_mean |> filter(
  Species == "ARTCAL"
)) |> summary()

artcal_chlorophyll

write_table(artcal_chlorophyll$coefficients, "anova-artcal", "chlorophyll")

salmel_chlorophyll = glm(Mean_Chlorophyll ~ Soil, data = chlorophyll_data_mean |> filter(
  Species == "SALMEL"
)) |> summary()

salmel_chlorophyll

write_table(salmel_chlorophyll$coefficients, "anova-salmel", "chlorophyll")


```
